---
title: "Class Work 2-1"
date: "Monday, September 23, 2019"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 1
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For Module 2, you will learn how to transform data using the package `dplyr`. `dplyr` is part of the `tidyverse` and provides a set of verbs for manipulating and transforming data.

*Note:* These class activities are adapted from "R for Data Science" by Grolemund and Wickham, chapter 5.  You can find this chapter [here](https://r4ds.had.co.nz/transform.html)

First we will load the `tidyverse` group of packages to have access to `dplyr`.

```{r}
library(tidyverse)
```

Remember that this loads the `tidyverse` group of packages; you can see that `dplyr` is one of these core packages.  We've already learned about `ggplot2` and we will learn about more of these later in the course.

We will again work with the National Medical Expenditures (NMES) data for this module, but we will work with the full set of data you used in Assignment 1.  Load the full `nmes.data` object into your workspace with the command below.

```{r}
load("nmes2018.rda")
```

Now back to our data! Recall that this `nmes.data` object is a data frame, which is a collection of observations (in the rows) across different variables (in the columns). Remember you can see the full dataset in a tab by typing:
```{r eval=FALSE}
View(nmes.data)
```

First lets look more closely at how the data prints when we view it in `R`:
```{r}
nmes.data
```

We only see the first few rows of the data and we only see the number of columns that can print on the screen.  Change your screen size and see again how it adapts!  This is because the data is stored in a **tibble**, which is a data structure that's easy to work with in the `tidyverse`.  We'll learn more about these later!  If we wanted to see the full data set (or as much of it as `R` will allow), we could view it as a data frame:

```{r eval=FALSE}
as.data.frame(nmes.data)
```

Which is easier to view?  When would you use one verses the other?

Going back to our tibble:
```{r}
nmes.data
```

Underneath each variable name is an abbreviation that tells us the type of the variable:

* `int` means integers
* `dbl` means doubles (or real numbers)
* `fct` means factors, how `R` represents categorical variables with fixed possible categories
* `chr` means characters (also called strings)

There are other types of variables not present in our dataset:

* `lgl` means logicals (values that can only be `TRUE` or `FALSE`)
* `date` means dates
* `dttm` means date-times (a date + a time)

The type of variable is important for what we are allowed to do with it; some of this is similar to the types of variables we have seen in class --
knowing whether a variable is continuous or binary helps us decide which type of summary to calculate (mean or proportion) or which type of graph to make (boxplot or barchart).

In this module we are going to learn 6 key functions to work with data: 

* `filter()` allows us to choose *observations* by their values; useful for defining subsets of the data that meet certain criteria
* `arrange()` allows us to order rows in the data based on some criteria
* `select()` allows us to choose *variables* from the dataset
* `mutate()` allows us to create new variables out of existing variables
* `summarize()` allows us to collapse values into a summary, such as collapsing individual values of a variable into the mean value or the sum of the values
* `group_by()` allows us to perform any of the previous 5 function on a group-by-group basis, for example summarizing to calculate a mean value for each group

All 6 of these function have a similar syntax:

* The first argument is a data frame (or tibble)
* The next arguments describe what to do with the data frame by referring to names of variables in the data set
* The result of the function is a new data frame

# Filter rows with `filter()`

The `filter()` function allows us to subsets our data by choosing observations (or rows) that meet certain criteria.  The first argument is the name of the data frame.  The second and following arguments are expressions that filter the data based on variable values.

**Example:** Only individuals who have ever smoked
```{r}
filter(nmes.data, eversmk=="Yes")
```

**Example:** Only individuals who spent more than $10,000 on medical expenditures
```{r}
filter(nmes.data, totalexp > 10000)
```

**Example:** Only never smokers who spent more than $10,000 on medical expenditures
```{r}
filter(nmes.data, eversmk=="No", totalexp > 10000)
```

If we want to save any of these *new* data frames for further use, we need to store the result as something using the *assignment operator*, `<-`:
```{r}
highSpenders <- filter(nmes.data, totalexp > 10000)
highSpenders
```

We can both store and print out the results in a single line by putting parentheses around the assignment command:
```{r}
(highSpenders <- filter(nmes.data, totalexp > 10000))
```

Notice that we filtered on 3 different conditions:

* `eversmk=="Yes"`
* `totalexp > 10000`
* `eversmk=="No"`

We define these conditions using comparisons: We compare the values in `eversmk` to either `"Yes"` or `"No"` and we compare the values in `totalexp` to the value 10000.  The standard comparison operators we can use are:

* `>` greater than
* `>=` greater than or equal to
* `<` less than
* `<=` less than or equal to
* `!=` not equal to
* `==` equal to
* `near()` equal to when using doubles, since the precision of the stored values might not match *exactly* out in the smallest digit!

```{r}
sqrt(2)^2
sqrt(2)^2 == 2
near(sqrt(2)^2, 2)

(1/49)*(7^2)
(1/49)*(7^2) == 1
near((1/49)*(7^2), 1)
```

Multiple arguments to the `filter()` function are combined with AND so that all of the expressions must be met in the filtered output.  So the following command subsets to individuals with `eversmk=="No"` AND `totalexp > 10000`.  This gives us never smokers with high expenditures:
```{r}
filter(nmes.data, eversmk=="No", totalexp > 10000)
```

We can also combine expressions with logical operators ourselves using:

* `&` means AND; all of the conditions must be met
* `|` means OR; at least one of the conditions must be met
* `!` means NOT; the condition must NOT be met

So we could also write the following to get never smokers with high expenditures:
```{r}
filter(nmes.data, eversmk=="No" & totalexp > 10000)
```

To get never smokers OR those with high expenditures we would write:
```{r}
filter(nmes.data, eversmk=="No" | totalexp > 10000)
```

To get all ever smokers EXCEPT those with high expenditures we could write:
```{r}
filter(nmes.data, eversmk=="Yes" & !(totalexp > 10000))
```

## Dealing with missing values

It's hard to do comparisons with missing values:
```{r}
NA > 5
NA == 10
NA == NA
```

We can't look at the subset with missing values for BMI just by filtering on `bwi == NA` because values for this vector will be `NA` when the condition is met instead of `TRUE`.  Try it:
```{r}
filter(nmes.data, bmi==NA)
```

Instead for missing data we use `is.na()` to decide whether a value is missing:
```{r}
filter(nmes.data, is.na(bmi))
```

So to exclude observations with missing values for BMI we could filter on `!is.na(bmi)`:
```{r}
filter(nmes.data, !is.na(bmi))
```

## Practice

1. Find all individuals who meet the following characteristics:

     a. Had more than $1,000 but less than $10,000 in medical expenditures
```{r}
     
```
     b. Always wear a seatbelt
```{r}
     
```
     c. Are obese (BMI >= 30) and currently smoke
```{r}
     
```
     d. Are both poor and college grads
```{r}
     
```
     e. Quit smoking more than 10 years ago
```{r}
     
```
     f. Quit smoking more than 10 years ago and are younger than 25 years old
```{r}
     
```
     g. Quit smoking more than 10 years ago or are younger than 25 years old
```{r}
     
```

2. Another useful filtering helper is `between()`.  Look at the help file for this function using `?between`.  Can you use it to simplify the code needed to answer the previous challenges?
```{r}
filter(nmes.data, totalexp > 1000, totalexp < 10000)
```

3. How many individuals are missing `bmi` values?
```{r}
```

4. How many individuals are missing `current` values?  Filter to these individuals.  Do you think this is really *missing* data?
```{r}

```


# Arrange rows with `arrange()`

The `arrange()` function changes the order of rows based on given criteria.  Again, the first argument is the name of the data frame.  The second and following arguments are either variable names or expressions containing variable names that are used to order the data set.  This function doesn't change the number of observations in the data set like `filter()` does, it just changes the order of the rows. 

```{r}
arrange(nmes.data, totalexp)
```

If more than one arrangement variable or expression is given, the later ones are used to break ties in the values of the preceeding orderings:
```{r}
arrange(nmes.data, totalexp, eversmk)
```

The default is to arrange in increasing (ascending) order.  To get decreasing (descending) order, use `desc()` within the `arrange()` function:
```{r}
arrange(nmes.data, desc(totalexp))
```

Missing values are always sorted to the end, regardless of whether you sort in increasing or descreasing order.

## Practice

1. Sort individuals in this dataset to find the youngest people. 
```{r}

```

2. Sort individuals in this dataset to find the oldest never smokers.
```{r}

```

3. Sort individuals in this dataset to find the oldest ever smokers
```{r}

```

4. Sort individuals in this dataset to find the never smokers with the highest medical expenditures.
```{r}

```

4. How could you use `arrange()` to sort all missing values of a variable to the start?  Hint: use `is.na()`.
```{r}

```

# Select columns with `select()`

The `select()` function allows you to select only a subset of the total variables in the dataset.  This can be especially useful if the dataset contains hundreds of different variables but you only need to work with a smaller number of them.  Again, the first argument to this function is the name of the data frame.  The second and following arguments are the variable names you want to keep, or expressions relating to the variable names you want to keep.

The `:` operator allows you to select all variables between two variables.  The `-` operator allows you to remove certain variables while keeping the rest.

```{r}
select(nmes.data, totalexp, eversmk)
select(nmes.data, totalexp:bmi)
select(nmes.data, -id)
select(nmes.data, -id, -lc5, -chd5)
```

There are helper functions to help make selecting variables easier, especially if variables share similar names.  These aren't that helpful for our NMES data, but may be when working with other data in the future:

* `starts_with("text")` matches names that begin with "text"
* `ends_with("text")` matches names that end with "text"
* `contains("text")` matches names that contain "text"
* `matches("(.)``1")` matches names that match a regular expression.  This one matches any variables that contain repeated characters.  You'll learn more about regular expression later in this course.
* `num_range("x", 1:3)` matches `x1`, `x2`, and `x3`.

So, for example, if we want to select both age variables, `age` and `ageCat`, we could do:
```{r}
select(nmes.data, starts_with("age"))
```

Or if we wanted both `packyears1` and `yearsince` and also `bmi` we could do:
```{r}
select(nmes.data, contains("years"), bmi)
```

We can also renames our variables when we select them:
```{r}
select(nmes.data, total_exp = totalexp, lung_cancer = lc5, ever_smoker = eversmk)
```

Notice that renaming variables with `select()` drops the other variables from the dataset.  If we just want to rename some variables but keep everything, we would use `rename()` instead of `select()`:
```{r}
rename(nmes.data, total_exp = totalexp, lung_cancer = lc5, ever_smoker = eversmk)
```

If you want to reorder the variables in your data frame to move the most important ones to the first columns but don't want to get rid of any variables, you can use `select()` with the `everything()` helper:
```{r}
select(nmes.data, totalexp, mscd, eversmk, bmi, age, everything())
```

And, again, remember that if you want to save your modified dataset so you can work with it, you need to assign the new dataset to an object with the assignment operator, `<-`:
```{r}
nmes.data <- select(nmes.data, totalexp, mscd, eversmk, bmi, age, everything())

nmes.data
```

## Practice

1. What happens if you include the name of a variable multiple times in a `select()` call?
```{r}
select(nmes.data, age, ageCat, bmi, bmi)
```

2. What does the `one_of()` function do? Run the code below to figure it out:
```{r}
myVars <- c("totalexp", "mscd", "age", "eversmk", "bmi")
select(nmes.data, one_of(myVars))
```


# Add new variables with `mutate()`

Sometimes we want to create a new variable from an old variable and add this new variable to our data set.  We can do this with the `mutate()` function.  Again, the first argument to this function is a data frame.  The following arguments define new variables in terms of variables already in the dataset:
```{r}
mutate(nmes.data, log10_exp = log10(totalexp))
```

If you look, you can see that a new variable called `log_exp` has been added to the end of the `nmes.data` data frame.  The `mutate()` function always adds variables to the end, so let's do that again after first creating a smaller data frame with fewer variables:
```{r}
nmes.sub <- select(nmes.data, totalexp, mscd, eversmk, bmi, age)

mutate(nmes.sub, log10_exp = log10(totalexp))
```

You can create more than one new variable at once, and can refer to columns you've previously created:
```{r}
mutate(nmes.sub, 
       log10_exp = log10(totalexp), 
       high_exp = (log10_exp > 4), 
       obese = (bmi >= 30)
       )
```

There are lots of different functions you can use together with `mutate()` to create new variables.  Here's a list of a bunch of them, but these are not all of them!

`+`, `-`, `*`, `/`, `^`, `sum()`, `mean()`, `log()`, `log10()`, `log2()`, `<`, `<=`, `>`, `>=`, `==`, `!=`, `min_rank()`, `percent_rank()`, and more!

For example, if you wanted to create a variable to give the percent rank of each individual in terms of medical expenditures:
```{r}
mutate(nmes.sub, rank_exp = percent_rank(totalexp))
```

## Practice

1. Create the following new variable and add to our NMES dataset in a single mutate command:

* Variable that tells whether an individual always uses their seat belt
* Variable that tells whether an individual's medical expenditures are higher than the mean expenditure value in the dataset
* Variable that tells whether an individual quit smoking more than 5 years ago

# Grouped summaries with `summarize()` and `group_by()`

The `summarize()` function condenses a data frame into a single row containing summary information calculated from the variables.  The first argument of this function is again a data frame and the following arguments name and define the summaries to be calculated from the data.

```{r}
summarize(nmes.data, mean_exp = mean(totalexp))
```

We can do multiple summaries at once:
```{r}
summarize(nmes.data, mean_exp = mean(totalexp), sd_exp = sd(totalexp), mean_bmi = mean(bmi), min_bmi = min(bmi), max_bmi = max(bmi))
```

Notice that we get `NA` as our summaries for the `bmi` variable.  This is because the `bmi` variable has missing values.  To get summaries that exclude the missing values, we can use the `na.rm = TRUE` option, in these functions:
```{r}
summarize(nmes.data, mean_exp = mean(totalexp), sd_exp = sd(totalexp), mean_bmi = mean(bmi, na.rm=TRUE), min_bmi = min(bmi, na.rm=TRUE), max_bmi = max(bmi, na.rm=TRUE))
```

The real power of the `summarize()` function happens when we use it in combination with the `group_by()` function.  This changes the summarization from happening to the full data set to happening on the group level instead.  See what happens when we first create a grouped dataset:
```{r}
by_smk <- group_by(nmes.data, eversmk)
summarize(by_smk, mean_exp = mean(totalexp), sd_exp = sd(totalexp), mean_bmi = mean(bmi, na.rm=TRUE), min_bmi = min(bmi, na.rm=TRUE), max_bmi = max(bmi, na.rm=TRUE))
```

The combination of `group_by()` and `summarize()` will be one of your major tools for working with data in `R`!

## Practice:

1. Calculate the mean medical expenditures for each of the age groups defined by `ageCat`.  What happens to medical expenditures with increasing age?
```{r}

```

2. Calculate the minimum and maximum ages for each of the education groups defined by `educate`.
```{r}

```

## Summarizing counts

Whenever we calculate a summary from our data, it's good practice to include the count of observations used in the summary along with the summary itself.  (You'll almost always see this, for example, in tables of summary statistics in journal articles!)  You can easily so this with the summarize function with `n()` to count the values or `sum(!is.na(x))` to count the non-missing values.  Check it out for mean medical expenditures by age:
```{r}
by_age <- group_by(nmes.data, ageCat)
summarize(by_age, n = n(), mean_exp = mean(totalexp))
```
For mean bmi values by age, we would need to count just the non-missing values:
```{r}
by_age <- group_by(nmes.data, ageCat)
summarize(by_age, n = sum(!is.na(bmi)), mean_bmi = mean(bmi, na.rm=TRUE))
```

We can see there are fewer individuals included in the mean BMI calculations than in the mean expenditure calculations.  This is important because we know that the accuracy of our estimates depend on sample size, so we want some idea of how many observations went into our summary calculation.

There are lots of different functions you can use together with `summarize()` to create meaningful summaries of your data.  Here's a list of a bunch of them, but these are not all of them!

* Measures of center: `mean()` and `median()`
* Measures of spread: `sd()`, `IQR()`, `range()`
* Measures of rank: `min()`, `max()`, `quantile()`
* Measures of position: `first()`, `nth()`, `last()`
* Counts: `n()`, `sum(!is.na())`, and `n_distinct()`
* Totals: `sum()`

## Grouping by multiple variables

We can group by more than one variable at a time, which allows us to calculate summaries for groups defined by combinations of variables:
```{r}
by_age_smk <- group_by(nmes.data, ageCat, eversmk)
summarize(by_age_smk, n = n(), mean_exp=mean(totalexp))
```

The ordering of the grouping is important for how the results are organized:
```{r}
by_age_smk <- group_by(nmes.data, ageCat, eversmk)
summarize(by_age_smk, n = n(), mean_exp=mean(totalexp))

by_smk_age <- group_by(nmes.data, eversmk, ageCat)
summarize(by_smk_age, n = n(), mean_exp=mean(totalexp))
```

The ordering is also important for taking the summarized data and doing another operation on it, which we we see in our next class!