---
title: "Class Work 4-2"
date: "Monday, November 4, 2019"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 1
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In Module 4, we are learning to write functions.  In Class 4-2, we will learn about function **arguments** and **return values** as well as **conditional execution**.

*Note:* These class activities are adapted from "R for Data Science" by Grolemund and Wickham, chapter 19.  You can find this chapter [here](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function)

# Recap from last week

There are three key steps to creating a new function using the following format:
```{r eval=FALSE}
function_name <- function(argument_1, argument_2, ...) {
# function's body
# code to do the function's task using the variables argument_1 and argument_2 from above
  
}
```

(1) You need to pick a **name** for the function. You should make the name meaningful to the task the function does.

(2) You list the inputs of the function, called **arguments**, inside `function()`. For a single arguemnt, we use `function(x)`.  For multiple arguments, we would have something like `function(x, y, z)`.  It can be helpful to give your arguments meaningful names as well!

(3) You place the code that does the task in the **body** of the function, a block between `{` and `}` that immediately follows `function(...)`.

**Example:** From Mini-Assignment 4-1, we wrote a function to calculate a confidence interval for a mean based on a sample from the population.
```{r}
mean_ci <- function(x) {
  xbar <- mean(x, na.rm=TRUE)
  n <- sum(!is.na(x))
  se <- sd(x, na.rm=TRUE) / sqrt(n)
  xbar + c(2, -2) * se
}

x <- runif(100)
mean_ci(x)
```

# Conditional execution

Sometimes when we write a function, we only want some of the code to run if a specific condition is met.  We can  conditionally execute code using an `if` statement:
```{r eval=FALSE}
if (condition) {
  # code executed when condition is TRUE
} else {
  # code executed when condition is FALSE
}
```

An example of what this might look like in our `mean_ci()` function from above:
```{r}
mean_ci <- function(x) {
  
  if (is.numeric(x)) {
    xbar <- mean(x, na.rm=TRUE)
    n <- sum(!is.na(x))
    se <- sd(x, na.rm=TRUE) / sqrt(n)
    xbar + c(2, -2) * se
  } else{
    "Cannot calculate CI since x is not a numeric variable."
  }
  
}
```

What is the `if` statement doing in this case?  We are using it to check whether the input to the function is numeric and we calculate a confidence interval from it.  If it's numeric, we calculate the confidence interval.  If it's not numeric, we print out a message telling the user what the problem is!

Try it:
```{r}
# using a numeric input
x <- runif(100)
mean_ci(x)

# using a non-numeric input
x <- c("Yes", "No", "Yes", "Yes", "No")
mean_ci(x)
```

This function demonstrates what's called the **standard return rule**: the function returns the last value that it computes.  In this case, the last thing computed depends on which brach of the `if` statement is used!

## Choosing a condition

The you use for the `condition` to the `if` statement must evaluate to either a single `TRUE` or single `FALSE` value.  If it doesn't, you'll either get a warning message or an error message: 
```{r eval=FALSE}
if (c(TRUE, FALSE)) {}

if (NA) {}

if(5) {}
```

You can combine multiple logical expressions using `||` for *or* and `&&` for *and*.  Notice that these are different that the symbols we used for *or* and *and* in the `dplyr` functions.  In `if` statements, you want to use the double `||` and `&&`.  

These would all be appropriate conditions to use:
```{r}
# if input is a single number
x <- 3.4

x > 5
x < 5 && x > 0
x > 5 || x < -5
is.numeric(x)
identical(x, 3.4)

# if input is a vector
x <- c(3.4, 5.3)

x > 5  # this one wouldn't work for a condition!  why?
any(x > 5)
all(x > 5)
any(!is.numeric(x))
any(is.na(x))
```

## Combining multiple conditions

You can chain multiple `if` statements together if you want to check a bunch of conditions in sequence:
```{r eval=FALSE}

if (this condition) {
  # do something
} else if (that condition) {
  # do something else
} else {
  # do a third thing if neither of previous conditions are met
}
```

Here's an example of what that might look like for a functions that combines multiple arithmetic operations:
```{r}
arith_xy <- function(x, y, operation) {
  if (operation == "plus") {
    x + y
  } else if (operation == "minus") {
    x - y
  } else if (operation == "times") {
    x * y
  } else if (operation == "divide") {
    x / y
  } else {
    "Unknown operation!  Please use either 'plus', 'minus', 'times', or 'divide'."
  }
}
```

We can try out this function:
```{r}
arith_xy(5, 3, "plus")
arith_xy(5, 3, "minus")
arith_xy(5, 3, "times")
arith_xy(5, 3, "divide")
arith_xy(5, 3, "subtract")
```

## Code format

In terms of code style, the `if` function should be followed by curly brackets (`{}`), and the contents should be indented by two spaces (RStudio does this automatically for you!) This makes it easier to see the hierarchy in your code by skimming the left-hand margin.

An opening curly brace `{` should never go on its own line and should always be followed by a new line. A closing curly brace  `}` should always go on its own line, unless it’s followed by `else`. Always indent the code inside curly braces.

```{r eval=FALSE}
# good example
if (y < 0) {
  message("Y is negative")
}

# bad example
if (y < 0)
message("Y is negative")

# good example
if (y == 0) {
  log(x)
} else {
  y ^ x
}

# bad example
if (y == 0) {
  log(x)
} 
else {
  y ^ x
}
```

## Practice

1. Write a `fizzbuzz` function. It takes a single number as input. If the number is divisible by three, it returns “fizz”. If it’s divisible by five it returns “buzz”. If it’s divisible by three and five, it returns “fizzbuzz”. Otherwise, it returns the number. Make sure you first write working code before you create the function.
```{r}

```

2. Write a `birthday` function that inputs a date and says “Happy birthday!” if it's your birthday and "X more days until your birthday" if it's not your birthday.
```{r}

```

# Function arguments

For functions that work with data, there are generally two types of arguemnts: ones that supply the data to comput on and ones that control the details of the computation to be done on that data.  For example:

* In `log()`, the data is `x`, and the detail is `base` which provides the base of the logarithm.
```{r}
?log
```

* In `mean()`, the data is `x`, and the details are `trim`, which tells how much data to trim from the ends and `na.rm` which tells how to handle missing values.
```{r}
?mean
```

* In `t.test()`, the data are `x` and `y`, and the details needed for the test are `alternative`, `mu`, `paired`, `var.equal`, and `conf.level`.
```{r}
?t.test
```

Generally, data arguments should come first. Detail arguments should go on the end, and usually should have default values. 

## Default values for detail arguments

You specify a default value by setting the default value in the argument portion of the `function()` function.  For example, see this updated function for calculating a confidence interval:
```{r}
mean_ci <- function(x, conf = 0.95) {
  xbar <- mean(x, na.rm=TRUE)
  n <- sum(!is.na(x))
  se <- sd(x, na.rm=TRUE) / sqrt(n)
  
  if (identical(conf, 0.95)) {
    xbar + c(2, -2) * se
  } else if (identical(conf, 0.99)) {
    xbar + c(2.6, -2.6) * se
  } else {
    "Confidence level not found; please use 0.95 or 0.99"
  }
}

x <- runif(100)
mean_ci(x)
mean_ci(x, conf = .95)
mean_ci(x, conf = .99)
mean_ci(x, conf = .9)
```

A good rule of thumb is to choose the default value for the details arguments to be the *most commonly used* value.  The one exception to this rule is when the most commonly used value might allow the user to not be aware of something important, such as the fact there are missing values.  So the default for `na.rm` arguments is usually `FALSE` so the user must be aware there are missing values and must make the choice to ignore them instead of ignoring them by default.

## Code style

When you call a function, you often omit the names of the data arguments because they are used so commonly. If you override the default value of a detail argument, you should use the full name:
```{r}
# Good
mean(1:10, na.rm = TRUE)

# Bad
mean(x = 1:10, , FALSE)
mean(, TRUE, x = c(1:10, NA))
```

Notice that when you call a function, you should place a space around = in function calls, and always put a space after a comma, not before (just like in regular English). Using whitespace makes it easier to skim the function for the important components.

```{r eval=FALSE}
# Good
average <- mean(feet / 12 + inches, na.rm = TRUE)

# Bad
average<-mean(feet/12+inches,na.rm=TRUE)
```

The names of the arguments are also important, because meaninful argument names are easier for the user! Generally you should use longer, more descriptive names, that make it clear what the argument does.  Also, it can be helpful to match names of your arguments to those in existing `R` functions.  For example, use `na.rm` to determine whether missing values should be removed!

## Checking values

Another useful thing to include in your functions are checks on whether the inputs to your function from the user are valid.  For example, in our `mean_ci()` function, we may want to be sure the vector `x` that is provided is numeric so that a mean CI is appropriate!  We can make this constraint explicit using the an `if` statement and then provide an error message to the user with the `stop()` function:

```{r error=TRUE}
mean_ci <- function(x, conf = 0.95) {
  if (!is.numeric(x)) {
    stop("`x` must be a numeric vector")
  }
  
  xbar <- mean(x, na.rm=TRUE)
  n <- sum(!is.na(x))
  se <- sd(x, na.rm=TRUE) / sqrt(n)
  
  if (identical(conf, 0.95)) {
    xbar + c(2, -2) * se
  } else if (identical(conf, 0.99)) {
    xbar + c(2.6, -2.6) * se
  } else {
    "Confidence level not found; please use 0.95 or 0.99"
  }
}

x <- runif(100)
mean_ci(x)

y <- c("hi", "there", "how", "are", "you", "?")
mean_ci(y)
```

We could also move the check of the confidence level to the top as well:
```{r error=TRUE}
mean_ci <- function(x, conf = 0.95) {
  if (!is.numeric(x)) {
    stop("`x` must be a numeric vector")
  }
  if (!identical(conf, 0.95) && !identical(conf, 0.99)) {
    stop("Confidence level not found; please use `0.95` or `0.99`")
  }
  
  xbar <- mean(x, na.rm=TRUE)
  n <- sum(!is.na(x))
  se <- sd(x, na.rm=TRUE) / sqrt(n)
  
  if (identical(conf, 0.95)) {
    xbar + c(2, -2) * se
  } else {
    xbar + c(2.6, -2.6) * se
  } 
}

x <- runif(100)
mean_ci(x)
mean_ci(x, conf = .95)
mean_ci(x, conf = .99)
mean_ci(x, conf = .9)
```


If there are a lot of conditions you want to check at the beginning of your function, you can use the functon `stopifnot()` that checks that each argument is TRUE, and produces a generic error message if not.  Note that when using `stopifnot()` you assert what should be TRUE rather than checking for what might be FALSE like we did in the `if` statement.
```{r error=TRUE}
mean_ci <- function(x, conf = 0.95) {
  stopifnot(is.numeric(x))
  stopifnot(identical(conf, 0.95) || identical(conf, 0.99))
  
  xbar <- mean(x, na.rm=TRUE)
  n <- sum(!is.na(x))
  se <- sd(x, na.rm=TRUE) / sqrt(n)
  
  if (identical(conf, 0.95)) {
    xbar + c(2, -2) * se
  } else {
    xbar + c(2.6, -2.6) * se
  } 
}

x <- runif(100)
mean_ci(x)
mean_ci(x, conf = .95)
mean_ci(x, conf = .99)
mean_ci(x, conf = .9)

y <- c("hi", "there", "how", "are", "you", "?")
mean_ci(y)
```


## Passing arguments through to other functions using `...`

Many functions in `R` can take an arbitrary number of inputs:
```{r}
sum(1, 2, 3, 4, 5)
sum(1, 2)
```

These functions rely on a special argument pronounced dot-dot-dot: `...`  You can see this special argument in the help file for these functions:
```{r}
?sum
```

This `...` argument captures any number of arguments that aren’t otherwise matched.  It’s a useful argument because you can then send those `...` arguments on to another function. This is a useful catch-all if your function includes other functions that have their own arguments that you might want to modify from within your function call. 

For example, in our original `mean_ci()` function, we used the `na.rm=TRUE` option for both the `mean()` and `sd()` functions that we used to calculate our confidence interval.  However, maybe we don't want the default to be `na.rm=TRUE` and instead want the user to be aware there is missing data!

We can use the `...` as an argument in our own function and then pass this `...` argument into the `mean()` and `sd()` functions we use in the body of our function:

```{r}
mean_ci <- function(x, ...) {
  xbar <- mean(x, ...)
  n <- sum(!is.na(x))
  se <- sd(x, ...) / sqrt(n)
  xbar + c(2, -2) * se
}

x <- runif(100)
mean_ci(x)

x <- c(runif(99), NA)
mean_ci(x)
mean_ci(x, na.rm=TRUE)
```

## Practice

1. Combine what we've learned so far to write a `mean_ci()` function that allows 90%, 95%, and 99% confidence intervals, checks to be sure the vector `x` is numeric, and allows the `na.rm` argument to pass through to the `mean()` and `sd()` functions.
```{r}

```

# Function return values

We already talked about the **standard return rule**, which means the function returns the last value that it computes.  We can also use explicit return statements to specify what we want the function to return.  We do this using the `return()` function.  This can be helpful if you want to be very clear about what value is being returned:
```{r}
mean_ci <- function(x, ...) {
  xbar <- mean(x, ...)
  n <- sum(!is.na(x))
  se <- sd(x, ...) / sqrt(n)
  ci <- xbar + c(2, -2) * se
  return(ci)
}

x <- runif(100)
mean_ci(x)
```

This can be helpful if you have different branches of an `if` statement:
```{r}
mean_ci <- function(x, conf = 0.95, ...) {
  stopifnot(is.numeric(x))
  stopifnot(identical(conf, 0.95) || identical(conf, 0.99))
  
  xbar <- mean(x, ...)
  n <- sum(!is.na(x))
  se <- sd(x, ...) / sqrt(n)
  
  if (identical(conf, 0.95)) {
    return(xbar + c(2, -2) * se)
  } else {
    return(xbar + c(2.6, -2.6) * se)
  } 
}

x <- runif(100)
mean_ci(x)
```



