---
title: "Class Work 2-2"
date: "Monday, September 30, 2019"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 1
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In Module 2, we are learning how to transform data using the package `dplyr`. `dplyr` is part of the `tidyverse` and provides a set of verbs for manipulating and transforming data.

*Note:* These class activities are adapted from "R for Data Science" by Grolemund and Wickham, chapter 5.  You can find this chapter [here](https://r4ds.had.co.nz/transform.html)

Load the `tidyverse` group of packages to have access to `dplyr` and the full NMES dataset:
```{r}
library(tidyverse)
load("nmes2018.rda")
```

In this module we are learning about 6 key functions from the `dplyr` package for working with data: 

* `filter()` allows us to choose *observations* by their values; useful for defining subsets of the data that meet certain criteria
* `arrange()` allows us to order rows in the data based on some criteria
* `select()` allows us to choose *variables* from the dataset
* `mutate()` allows us to create new variables out of existing variables
* `summarize()` allows us to collapse values into a summary, such as collapsing individual values of a variable into the mean value or the sum of the values
* `group_by()` allows us to perform any of the previous 5 function on a group-by-group basis, for example summarizing to calculate a mean value for each group

A reminder that all 6 of these function have a similar syntax:

* The first argument is a data frame (or tibble)
* The next arguments describe what to do with the data frame by referring to names of variables in the data set
* The result of the function is a new data frame

We talked about `filter()`, `arrange()`, `select()`, and `mutate()` last week.  This week we will look at `group_by()` and `summarize()` and learn how to work with these functions in combination!

# Grouped summaries with `summarize()` and `group_by()`

The `summarize()` function condenses a data frame into a single row containing summary information calculated from the variables.  The first argument of this function is again a data frame and the following arguments name and define the summaries to be calculated from the data.

```{r}
summarize(nmes.data, mean_exp = mean(totalexp))
```

We can do multiple summaries at once:
```{r}
summarize(nmes.data, mean_exp = mean(totalexp), sd_exp = sd(totalexp), mean_bmi = mean(bmi), min_bmi = min(bmi), max_bmi = max(bmi))
```

Notice that we get `NA` as our summaries for the `bmi` variable.  This is because the `bmi` variable has missing values.  To get summaries that exclude the missing values, we can use the `na.rm = TRUE` option, in these functions:
```{r}
summarize(nmes.data, mean_exp = mean(totalexp), sd_exp = sd(totalexp), mean_bmi = mean(bmi, na.rm=TRUE), min_bmi = min(bmi, na.rm=TRUE), max_bmi = max(bmi, na.rm=TRUE))
```

The real power of the `summarize()` function happens when we use it in combination with the `group_by()` function.  This changes the summarization from happening to the full data set to happening on the group level instead.  See what happens when we first create a grouped dataset:
```{r}
by_smk <- group_by(nmes.data, eversmk)
summarize(by_smk, mean_exp = mean(totalexp), sd_exp = sd(totalexp), mean_bmi = mean(bmi, na.rm=TRUE), min_bmi = min(bmi, na.rm=TRUE), max_bmi = max(bmi, na.rm=TRUE))
```

The combination of `group_by()` and `summarize()` will be one of your major tools for working with data in `R`!

## Practice:

1. Calculate the mean medical expenditures for each of the age groups defined by `ageCat`.  What happens to medical expenditures with increasing age?
```{r}
by_ageGroup <- group_by(nmes.data, ageCat)
summarize(by_ageGroup, mean_exp = mean(totalexp))
```

2. Calculate the minimum and maximum ages for each of the education groups defined by `educate`.
```{r}
by_edu <- group_by(nmes.data, educate)
summarize(by_edu, min_age = min(age), max_age = max(age))
```

## Summarizing counts

Whenever we calculate a summary from our data, it's good practice to include the count of observations used in the summary along with the summary itself.  (You'll almost always see this, for example, in tables of summary statistics in journal articles!)  You can easily so this with the summarize function with `n()` to count the values or `sum(!is.na(x))` to count the non-missing values.  Check it out for mean medical expenditures by age:
```{r}
by_age <- group_by(nmes.data, ageCat)
summarize(by_age, n = n(), mean_exp = mean(totalexp))
```
For mean bmi values by age, we would need to count just the non-missing values:
```{r}
by_age <- group_by(nmes.data, ageCat)
summarize(by_age, n = sum(!is.na(bmi)), mean_bmi = mean(bmi, na.rm=TRUE))
```

We can see there are fewer individuals included in the mean BMI calculations than in the mean expenditure calculations.  This is important because we know that the accuracy of our estimates depend on sample size, so we want some idea of how many observations went into our summary calculation.

There are lots of different functions you can use together with `summarize()` to create meaningful summaries of your data.  Here's a list of a bunch of them, but these are not all of them!

* Measures of center: `mean()` and `median()`
* Measures of spread: `sd()`, `IQR()`, `range()`
* Measures of rank: `min()`, `max()`, `quantile()`
* Measures of position: `first()`, `nth()`, `last()`
* Counts: `n()`, `sum(!is.na())`, and `n_distinct()`
* Totals: `sum()`

## Grouping by multiple variables

We can group by more than one variable at a time, which allows us to calculate summaries for groups defined by combinations of variables:
```{r}
by_age_smk <- group_by(nmes.data, ageCat, eversmk)
summarize(by_age_smk, n = n(), mean_exp=mean(totalexp))
```

The ordering of the grouping is important for how the results are organized:
```{r}
by_age_smk <- group_by(nmes.data, ageCat, eversmk)
summarize(by_age_smk, n = n(), mean_exp=mean(totalexp))

by_smk_age <- group_by(nmes.data, eversmk, ageCat)
summarize(by_smk_age, n = n(), mean_exp=mean(totalexp))
```

The ordering is also important for taking the summarized data and doing another operation on it, which we will see later!

## Practice

1. Find the number of people, mean, and median medical expenditures for those who have and don't have a major smoking-caused disease (`mscd`).
```{r}
by_mscd <- group_by(nmes.data, mscd)
summarize(by_mscd, n=n(), mean_exp=mean(totalexp), med_exp=median(totalexp))
```

2. Find the number of people, mean, and median BMIs for those who have and haven't ever smoked (`eversmk`).
```{r}

```

# Combining multiple operations with the pipe, `%>%`

As you can see with `group_by()` and `summarize()`, sometimes we want to perform different data operations in sequence -- or one after another.  We could think about adding even more operations to this sequence: 

**If we wanted to compare mean medical expenditures between age groups but only for those who are not obese we would:**

1. Filter `nmes.data` to only the non-obese individuals; store this new data frame as `nmes.notObese`:
```{r}
nmes.notObese <- filter(nmes.data, bmi < 30)
```

2. Group the `nmes.notObese` data by age categories; store this new grouped data frame as `nmes.notObese.by_age`:
```{r}
nmes.notObese.by_age <- group_by(nmes.notObese, ageCat)
```

3. Summarize the `nmes.notObese.by_age` data frame by calculating the mean expenditures:
```{r}
summarize(nmes.notObese.by_age, mean_exp=mean(totalexp))
```

To do this, we need to save our modified data frame each time we change it, even though all we really care about is the very last result and not the intermediate steps.  And you can see how our data frame names could become out of control if we had to do many more steps!

A better option is to use the pipe, designated by `%>%` to take the result from one of these operations and "pipe" it straight into the next one without saving the intermediate results. Basically this pipe operation takes the result from the thing on the left and puts it into the first argument of the thing on the right.  Since each of these operations take a data frame as the first argument and return a modified data frame as a result, this works perfectly for what we want!

```{r}
nmes.data %>%
  filter(bmi < 30) %>%
  group_by(ageCat) %>%
  summarize(mean_exp = mean(totalexp))
```

Writing our code this way is a lot cleaner and actually makes what we are doing more clear!  If you think about reading the `%>%` symbol as "then", you might read this code as:

Start with `nmes.data`. **Then** filter to `bmi < 30`. **Then** group by `ageCat`. **Then** summarize by calculating mean `totalexp`.

## Practice

Use the pipe to perform the following sequence of operations on the NMES data: create a log10 medical expenditures variable, filter to only people older than 65, group by smoking status, summarize to get median log10 expenditures and the number of people in each smoking group
```{r}
nmes.data %>%
  mutate(log10_exp = log10(totalexp)) %>%
  filter(age > 65) %>%
  group_by(eversmk) %>%
  summarize(n=n(), med_log10exp=median(log10_exp))
```

## Back to grouping by multiple variables

Recall that we can group by more than one variable at a time.  This is what it would look like using the pipe:
```{r}
nmes.data %>%
  group_by(ageCat, eversmk) %>%
  summarize(n = n(), mean_exp=mean(totalexp))
```

Again, recall that the ordering of the grouping is important for how the results are organized:
```{r}
nmes.data %>%
  group_by(ageCat, eversmk) %>%
  summarize(n = n(), mean_exp=mean(totalexp))

nmes.data %>%
  group_by(eversmk, ageCat) %>%
  summarize(n = n(), mean_exp=mean(totalexp))
```

The ordering is also important for taking the summarized data and doing another operation on it.  Suppose we want to calculate the proportion within each age group who are ever smokers. We might first group by `ageCat` and `eversmk` to get the number of ever smokers and never smokers in each age group:
```{r}
nmes.data %>%
  group_by(ageCat, eversmk) %>%
  summarize(n = n())
```

Then we could mutate these results to get a proportion by taking the counts over the sum of the counts for that group:
```{r}
nmes.data %>%
  group_by(ageCat, eversmk) %>%
  summarize(n = n()) %>%
  mutate(prop = n/sum(n))
```
Notice here that the proportions add up to 1 within an age group; this is because the the `sum(n)` in the mutate command above will work on the last group defined by `group_by()`, so will sum across the `eversmk` groups.  This correctly gives us the proportion of ever smokers and never smokers within each age group.

However, if we had grouped our variables in the different order, we would get something different!
```{r}
nmes.data %>%
  group_by(eversmk, ageCat) %>%
  summarize(n = n()) %>%
  mutate(prop = n/sum(n))
```

Now we see that the proportions add up to 1 within a smoking group because the last group defined is `ageCat` so the denominator in our proportion is summing across the age groups.   So here we've really calculated the proportion in each age Group within a smoking group, which is not quite what we wanted to get.

## Practice

1. Use the pipe to calculate the proportions of people with and without MSCD within each smoking group.
```{r}

```

2. Modify your code from (1) to only give the proportions *with* MSCD for each smoking group.
```{r}

```


## Ungrouping

If you need to remove the grouping, you can use the `ungroup()` function:
```{r}
nmes.data %>%
  group_by(eversmk, ageCat) %>%
  summarize(n = n()) %>%
  mutate(prop = n/sum(n)) %>%
  ungroup()
```

# Combining `dplyr` with `ggplot2`

Since the first argument in the `ggplot()` function is a data frame, we can pipe a data frame directly into the `ggplot()` function to create graphs:
```{r}
nmes.data %>%
  ggplot(mapping = aes(x = age, y = totalexp)) +
  geom_point() + 
  geom_smooth()
```

Notice two things:

(1) We don't have to give a `data=` option in `ggplot()` because the data is coming through the pipe.
(2) We still use plus signs (`+`) to join layers of the graphic together.

This options is incredibly useful if we first want to modify our data using the `dplyr` functions before creating the graph.  For example, if we want to create and graph a new variable:
```{r}
nmes.data %>%
  mutate(log10exp = log10(totalexp)) %>%
  ggplot(mapping = aes(x=age, y=log10exp)) +
  geom_point() + 
  geom_smooth(se=FALSE)
```

Or if we want to filter to only a certain group of individuals:
```{r}
nmes.data %>%
  filter(eversmk=="Yes") %>%
  ggplot(mapping = aes(x=age, y=totalexp)) +
  geom_point() + 
  geom_smooth(se=FALSE)
```

Or both:
```{r}
nmes.data %>%
  filter(eversmk=="Yes") %>%
  mutate(log10exp = log10(totalexp)) %>%
  ggplot(mapping = aes(x=age, y=log10exp)) +
  geom_point() + 
  geom_smooth(se=FALSE)
```


This works well for all types of graphs, such as barplots:
```{r}
nmes.data %>%
  group_by(ageCat, eversmk) %>%
  summarize(n = n()) %>%
  mutate(proportion = n/sum(n))

nmes.data %>%
  group_by(ageCat, eversmk) %>%
  summarize(n = n()) %>%
  mutate(proportion = n/sum(n)) %>%
  ggplot(mapping = aes(x = ageCat, fill = eversmk)) +
  geom_bar(mapping = aes(x = ageCat, y = proportion, fill = eversmk), stat="identity", position="dodge")

nmes.data %>%
  group_by(ageCat, eversmk) %>%
  summarize(n = n()) %>%
  mutate(proportion = n/sum(n)) %>%
  filter(eversmk=="Yes") %>%
  ggplot(mapping = aes(x = ageCat, fill = eversmk)) +
  geom_bar(mapping = aes(x = ageCat, y = proportion, fill = eversmk), stat="identity", position="dodge")
```