---
title: "Class Work 4-3"
date: "Monday, November 11, 2019"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 1
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In Module 4, we are learning to write functions.  In Class 4-3, we will practice writing functions.

*Note:* Our class materials for module 4 are adapted from "R for Data Science" by Grolemund and Wickham, chapter 19.  You can find this chapter [here](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function)

# Some key information from previous classes

We create a new function using the following format:
```{r eval=FALSE}
function_name <- function(argument_1, argument_2, ...) {
# function's body
# code to do the function's task using the variables argument_1 and argument_2 from above
  
}
```

We can specify a **default value** by setting the default value in the argument portion of the `function()` function, such as `function(argument_1, argument_2 = default)`.

We can use condition execution if we want to run code only if a specific condition is met:
```{r eval=FALSE}
if (condition) {
  # code executed when condition is TRUE
} else {
  # code executed when condition is FALSE
}
```

We can chain multiple `if` statements together if we want to check a bunch of conditions in sequence:
```{r eval=FALSE}

if (this condition) {
  # do something
} else if (that condition) {
  # do something else
} else {
  # do a third thing if neither of previous conditions are met
}
```

We can **check** whether the values given to the function are valid using an `if` statement and then providing an error message to the user with the `stop()` function:
```{r eval=FALSE}
  if (condition) {
    stop("error message")
  }
```

We can pass arguments (like `na.rm = TRUE`) through our function to embedded functions using the `...` argument in our function:
```{r eval=FALSE}
function_name <- function(argument_1, argument_2, ...) {

  mean(argument1, ...)
  
}
```

The **standard return rule** says that a function returns the last value that it computes.  We can force a value to return early by using `return(value)` within our function.

# Practice

1. Write a function `fahrenheit_to_celsius()` that converts temperatures from Fahrenheit to Celsius.
```{r}
# x <- 63  # temp in F
# (x - 32) * 5/9

f_to_c <- function(x) {
  # check that input is numeric
  if(!is.numeric(x)) {   # TRUE if not numeric, FALSE if numeric
    stop("x must be numeric")
  }
  
  (x - 32) * 5/9 
}

f_to_c(63)
f_to_c(95)
f_to_c(212) 
f_to_c("five")
f_to_c(five)

# some condition to check on the input -- make sure the input value is numeric, no restrictions beyond that

# another way

f_to_c <- function(x) {
  # check that input is numeric
  if(is.numeric(x)) {   # TRUE if  numeric, FALSE if not numeric
    (x - 32) * 5/9 
  } else {  
    stop("x must be numeric")
  }
}

```

2. Write a `temperature_conversion()` function that will convert temperatures from Fahrenheit to Celsius OR from Celsius to Fahrenheit, depending on what the user specifies.
```{r}
#example inputs
x <- 23
conv_type <- "f2c"

(x - 32) * 5/9 

#other example inputs
x <- 23
conv_type <- "c2f"

(x * 9/5) + 32

# build function

temp_conversion <- function(x, conv_type = "f2c") {  # we will make f2c the default convergence type
    # check that input is numeric
  if(!is.numeric(x)) {   # TRUE if not numeric, FALSE if numeric
    stop("x must be numeric")
  }
  
  if (conv_type == "f2c") {
    (x - 32) * 5/9 
  } else if (conv_type == "c2f") {
    (x * 9/5) + 32
  } else {
    "Please specify conv_type as `f2c` or `c2f`."
  }

}

temp_conversion(23)
temp_conversion(23, conv_type="f2c")
temp_conversion(23, conv_type="c2f")
temp_conversion("five", conv_type="c2f")

# we could add a check at the beginning to check if x is numeric
```

3. Write a `logit()` function that calculates the log odds, log(p/(1-p)) of an input.  Be sure to check that the input is a valid probability between 0 and 1.
```{r}
p <- c(.5, .3, 1.3)
log( p / (1-p) )

logit <- function(p) {
  # check that p is between 0 and 1
  if ( any(p < 0) || any(p > 1)) {   # give the condition on input that will break my function -- give the BAD, checks all vals in vector
    stop("p must contain values between 0 and 1")
  }
  log( p / (1-p) ) 
}

logit(.3)
logit(.7)
logit(1.4)
```

4. Write a `birthday` function that inputs a date in the format `"mm-dd"` and says “Happy birthday!” if it's your birthday and "It's not your birthday yet!" if it's not your birthday.
```{r}

```

5. Write a function called `prop_ci()` that takes the arguments listed below and returns either a 95% confidence interval for a single proportion or a 95% confidence interval for the difference in proportions depending on those arguments.  I've given default values for these arguments as well.

* `x` (no default value)
* `y` (default value is `NULL` which corresponds to `one.prop` option for the `type` argument)
* `type` (default value is `one.prop`, other value is `two.props`)
* `...` (to be able to pass the `na.rm` option through to any functions as needed)

Consider including a a way to check whether the vectors of `x` and `y` are 0/1 and thus such a confidence interval is appropriate.

Hint: To be able to use any confidence level between 0 and 1, note that to get the multiplier for a 95% confidence interval. we could use the code: `qnorm(.975)`. To get the multiplier for a 99% confidence interval, we could use the code: `qnorm(.995)`.  To get the multiplier for a 90% confidence interval, we could use the code `qnorm(.95)`.  Can you generalize this to any confidence level? **not needed anymore**
```{r}
## type = "one.props"
x <- c(0,1,1,0,1)
y <- NULL
type <- "one.props"

  if ( any(x != 1 & x != 0) ) {   
    stop("Please use data that is only 0/1.")
  }

# build the code for this case
# look at solutions to mini-assignment 4-1



## type = "two.props"
x <- c(0,1,1,0,1)
y <- c(1,1,1,0,0,0,0)
type <- "two.props"

# build the code for this case
# look at solutions to mini-assignment 4-1


```



